Conductive and convective heat transfer in inductive heating of subsea buried pipelines

### 2021-10-08 manuscript.tex
:pencil: Complete draft of the paper

\begin{equation}
Q_{in} = - \lambda \nabla^2 T + \rho_f c_{p_f} \mathbf{u} \cdot \nabla T\,,
\label{eq:pde-heat}
\end{equation}
\noindent where $T$ is Temperature (\si{\celsius}), $\rho_f$ is the density of fluid (\si{\kilo\gram\per\meter\cubed}), $c_{p_f}$ is the specific heat capacity of fluid (\si{\joule\per\kilo\gram\per\celsius}), and $\mathbf{u}$ is the fluid velocity (\si{\meter\per\second}). The right hand side of~\cref{eq:pde-heat} represents the conductive ($- \lambda \nabla^2 T$) and convective ($ \rho_f c_{p_f} \mathbf{u} \cdot \nabla T$) heat transfer mechanisms. The bulk thermal conductivity of the soil $\lambda$ controls the conductive heat transfer. The bulk thermal conductivity is a function of porosity ($n$) and the individual thermal conductivities of solid ($\lambda_s$) and fluid ($\lambda_f$):

@@

\Cref{tab:mat} presents the thermal properties of the fluid, soil, and cables. The bulk thermal conductivity and bulk specific heat of the porous media are calculated using simple arithmetic based on the volume fraction of fluid and solids:

@@

\begin{table}[htbp]
\caption{Thermal properties of the fluid, soil, and the cables}
\label{tab:mat}
\centering
\begin{tabular}{ll}
\toprule
\textbf{Properties}                         & \textbf{Value}   \\
\midrule
\textbf{Fluid}                              &                  \\
Fluid density ($\rho_f$)                    & 1000 \si{\kilo\gram\per\meter\cubed}    \\
Specific heat of fluid ($Cp_f$)             & 4290 \si{\joule\per\kilogram-\kelvin}    \\
Thermal conductivity of fluid ($\lambda_f$) & 0.6 \si{\watt\per\meter-\kelvin} \\
Thermal diffusivity of fluid ($\alpha_f$)   & 1.39E-7 \si{\meter\squared\per\second}  \\
Dynamic viscosity of fluid ($\mu$)          & 1E-3 \si{\pascal\cdot\second}      \\
\textbf{Soil}                               &                  \\
Specific heat of soil ($C_{p_s}$)           & 8000 $J/kg-K$    \\
Thermal conductivity of soil ($\lambda_s$)  & 1.0 $W/m-K$      \\
\textbf{Pipe}                               &                  \\
Thermal diffusivity of pipe ($\alpha_p$)    & 1.172E-5 $m^2/s$ \\
\bottomrule
\end{tabular}
\end{table}

@@

\Cref{fig:dTdt} shows the rate of temperature change with time at the center of the pipe buried in soils with different permeabilities to identify the ideal simulation duration. For high permeability soils ($k > \SI{1E-10}{\meter\squared}$), the steady-state condition is reached within 50 hours, whereas low permeability soils ($K < \SI{1E-12}{\meter\squared}$) show no noticeable change in temperature gradients beyond 150 hours.~\Cref{fig:200,fig:500,fig:1000} shows the heat distribution remains unchanged beyond 200 hours. We choose 200 hours for all subsequent simulations to balance the long computational duration and achieve a system closer to steady-state.

@@

% \pagebreak
% %
% \appendix
% %
% \section{Notation}
% \label{app:notation}
% The following symbols are used in this paper:
% \nopagebreak
% \par
% \begin{tabular}{r  @{\hspace{1em}=\hspace{1em}}  l}
% $D$                    & Depth of burial (m); \\
% \end{tabular}

% \label{section:references}
\bibliography{ascexmpl-new}

>>>>

\begin{equation}
Q_{in} = - \lambda \nabla^2 T + \rho_f c_{p_f} \mathbf{u} \cdot \nabla T\,,
\label{eq:pde-heat}
\end{equation}
where $\lambda$ is the thermal conductivity (\si{\watt\per\meter-\kelvin}), $T$ is Temperature (\si{\celsius}), $\rho_f$ is the density of fluid (\si{\kilo\gram\per\meter\cubed}), $c_{p_f}$ is the specific heat capacity of fluid (\si{\joule\per\kilo\gram\per\celsius}), and $\mathbf{u}$ is the fluid velocity (\si{\meter\per\second}). The right hand side of~\cref{eq:pde-heat} represents the conductive ($- \lambda \nabla^2 T$) and convective ($ \rho_f c_{p_f} \mathbf{u} \cdot \nabla T$) heat transfer mechanisms. The bulk thermal conductivity of the soil $\lambda$ controls the conductive heat transfer. The bulk thermal conductivity is a function of porosity ($n$) and the individual thermal conductivities of solid ($\lambda_s$) and fluid ($\lambda_f$):

@@

\Cref{tab:mat} presents the thermal properties of the fluid, soil, and cables. The thermal conductivity of soil depends on the water content~\cite{young2001utilizing}. The bulk thermal conductivity and bulk specific heat of the porous media are calculated using simple arithmetic based on the volume fraction of fluid and solids:

@@

\begin{table}[htbp]
\caption{Thermal properties of the fluid, soil, and the cables}
\label{tab:mat}
\centering
\begin{tabular}{ll}
\toprule
\textbf{Properties}                         & \textbf{Value}   \\
\midrule
\textbf{Fluid}                              &                  \\
Fluid density ($\rho_f$)                    & 1000 \si{\kilo\gram\per\meter\cubed}    \\
Specific heat of fluid ($Cp_f$)             & 4290 \si{\joule\per\kilogram-\kelvin}    \\
Thermal conductivity of fluid ($\lambda_f$) & 0.6 \si{\watt\per\meter-\kelvin} \\
Thermal diffusivity of fluid ($\alpha_f$)   & 1.39E-7 \si{\meter\squared\per\second}  \\
Dynamic viscosity of fluid ($\mu$)          & 1E-3 \si{\pascal\cdot\second}      \\
\textbf{Soil}                               &                  \\
Specific heat of soil ($C_{p_s}$)           & 8000 \si{\joule\per \kilogram-\kelvin}    \\
Thermal conductivity of soil ($\lambda_s$)  & 1.0 \si{\watt\per\meter-\kelvin}      \\
\textbf{Pipe}                               &                  \\
Thermal diffusivity of pipe ($\alpha_p$)    & 1.172E-5 \si{\meter\squared\per\second}  \\
\bottomrule
\end{tabular}
\end{table}

@@

\Cref{fig:dTdt} shows the rate of temperature change with time at the center of the pipe buried in soils with different permeabilities to identify the ideal simulation duration. For high permeability soils ($k > \SI{1E-10}{\meter\squared}$), the steady-state condition is reached within 50 hours, whereas low permeability soils ($k < \SI{1E-12}{\meter\squared}$) show no noticeable change in temperature gradients beyond 150 hours.~\Cref{fig:200,fig:500,fig:1000} shows the heat distribution remains unchanged beyond 200 hours. We choose 200 hours for all subsequent simulations to balance the long computational duration and achieve a system closer to steady-state.

@@

The heat transfer mechanism changes from conduction to convection as the permeability of soil increases. A critical Rayleigh number defines this transition point from conduction to convection.~\citeN{nield1968onset} defines this transition at a critical $Ra$ of 12. At low permeabilities, where $Ra < 1$, the temperature field is symmetric, indicating a conductive behavior (see~\cref{fig:ra-evol}). The asymmetric convective behavior occurs for $Ra > 2.22$ ($k > \SI{4.84E-12}{\meter\squared}$). The observation of convective response for $Ra = 2.22$, below the critical $Ra$ of 12 for a rectangular domain with Free-Free boundary condition, indicates the transition occurs earlier than originally thought.

\Cref{fig:ra-vline} shows the temperature distribution with depth along the A-A' axis for different Rayleigh numbers. The fluid temperature in the pipe reaches a maximum of 73.5\% of the cable temperature for the lowest $Ra = 4.5E-4$ ($k = \SI{1.18E-15}{\meter\squared})$. The fluid temperature inside the pipe remains almost unchanged with increasing Rayleigh number until a $Ra = 0.46$ ($k > \SI{1E-12}{\meter\squared}$), above which the fluid temperature drops drastically. For large Rayleigh numbers ($Ra > 100$), the fluid in the pipe barely reaches 5\% of the cable temperature. Below the pipe, the heat dissipation profile remains unchanged with increasing $Ra$, which indicates conductive behavior dominating the region below the pipe. The heat distribution above the cable varies as the Rayleigh number increases due to buoyancy forces. Lower $Ra$ conditions show a lower rate of dissipation (concave profile), whereas soils experiencing a higher Rayleigh number show an accelerated heat distribution (convex profile). The heat profile above the cables transitions from concave to convex between a Rayleigh number of 2 and 7, below the established critical Ra of 12~\cite{nield1968onset}.

We closely examine the heat distribution above and below the pipeline to determine different modes of heat transfer.~\Cref{fig:heat-pipe} shows the evolution of temperature in the pipe and above the cables for different Rayleigh numbers. For $Ra < 1$, the average pipe temperature remains unchanged at 73.5\% of the cable temperature, and the soil temperature above the cable stays constant at 28\% of the cable temperature. As the Rayleigh number increases from 1 to 100, the temperature in the pipe drastically drops to 13\%, while the soil temperature rises to 95\% of cable temperature. The heating of pipe using cables is nonexistent for $Ra > 100$, where convection transfers almost all of the cable heat to the soil above. Hence, the transition from conduction to convection happens below a Rayleigh number of 1.0, much earlier than previously thought.

\begin{figure}
    \centering
    \includegraphics{figs/temp-pipe.eps}
    \caption{Temperature distribution along the vertical axis (A-A’) through the center of pipe for different Rayleigh numbers.}
    \label{fig:ra-vline}
\end{figure}

\begin{figure}
    \centering
    \includegraphics{figs/heat-pipe.eps}
    \caption{The average temperature in pipes buried in soils with different Rayleigh numbers.}
    \label{fig:heat-pipe}
\end{figure}

In addition to soil permeability, the temperature in the cable also influences the Rayleigh-Darcy number, and in turn, the transition from conduction to convection. We vary the temperature in the cable from \SI{5}{\celsius} to \SI{90}{\celsius} for different soil permeabilities.~\Cref{fig:k-temp} summarizes the combined influence of temperature and permeability on the pipe temperature. We use the normalized pipe temperature ($T_{pipe}/T_{cable}$) to investigate how the cable temperature influences the heat transfer mechanism at different permeabilities. For low permeabilities ($k < \SI{1E-12}{\meter\squared}$), irrespective of the applied cable temperature, the pipe temperature reaches a constant maximum of 73.5\% of the cable temperature. The constant pipe temperature in low permeability soil indicates that conduction is the primary heat transfer mechanism. A transition region exists for a permeability range of \SI{1E-12} to \SI{1E-9}{\meter\squared}, where conduction and convection play a role in distributing the heat from the cables. In the transition zone, the efficiency in heating the pipe decreases as the permeability increases—the proportion of convection to conduction controls the heat transfer efficiency. For a given permeability, the lower the temperature higher is the efficiency of heat transfer to the pipes. In other words, as the cable temperature increases, more heat is lost to the ground surface due to convection, and the heat transferred to the pipe drops. An increase in cable temperature increases the Rayleigh number proportionally (see~\cref{eq:ra}), thus favoring convection over conduction. As the permeability increases beyond \SI{1E-9}{\meter\squared}, only convection occurs, driving the heat away from the cable towards the ground surface, thus barely heating the pipe irrespective of cable temperature. Hence, the Rayleigh number, and not permeability, is the controlling factor in determining the mode of heat transfer.

\begin{figure}
    \centering
    \includegraphics{figs/k-dt.eps}
    \caption{Normalized pipe temperature for different soil permeabilities.}
    \label{fig:k-temp}
\end{figure}

We replot the normalized cable temperature for different soil permeabilities as a function of the Rayleigh number (see~\cref{fig:ra-temp}) to show the dependence of heat transfer on the Rayleigh number. The normalized pipe temperature across all permeabilities collapses onto a single curve. The normalized pipe temperature remains constant when conduction dominates ($Ra < 1$). In the conduction region, the heat transfer has a maximum efficiency of 73.5\% for a burial depth of \SI{0.5}{\meter}. As the Rayleigh number increases between 1 and 100, the efficiency of heat transfer drops drastically. An increase in the Rayleigh number means an increase in convection over conduction, driving heat away from the cables towards the ground surface. Convection is the dominant heat transfer mechanism at higher Rayleigh numbers ($Ra > 100$) and has the lowest heat transfer efficiency.

\begin{figure}
    \centering
    \includegraphics{figs/ra-dt.eps}
    \caption{Normalized pipe temperature for different Rayleigh numbers.}
    \label{fig:ra-temp}
\end{figure}

\Cref{fig:ra-temp-contour} shows the variation in the pipe temperature for different soil permeabilities and cable temperatures. Conduction predominates for low permeability soils ($k < \SI{1E-12}{\meter\squared}$). With negligible convection, there is no heat loss away from the cables. Conversely, for very high permeability soils ($k > \SI{1E-10}{\meter\squared}$), essentially all the heat is transferred away from the cables by convection. The onset of convective behavior also depends on the cable temperature. Convection increases with an increase in the cable temperature. We identify a transition at a $Ra$ of 1, which separates regions dominated by conduction ($Ra < 1$) and convection ($Ra > 1$). Above an $Ra$ of 100, only convective flow occurs, and the pipe remains unheated.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figs/conduction-convection.png}
    \caption{Variation of the pipe temperature for different soil permeabilities and cable temperatures. The solid curves show different Rayleigh numbers increase from left to right with an increase in permeability.}
    \label{fig:ra-temp-contour}
\end{figure}

\subsection*{Relationship between heat transfer and flow condition}
The Rayleigh number ($Ra$) defines the amount of free convective flow, but not the amount of heat transfer. Whereas, the Nusselt number ($Nu$) defines the ratio of convective and conductive heat transfer at a boundary. While an increase in flow increases the heat exchange, the relationship between the Nusselt number and the Rayleigh number is not necessarily linear. In regions of natural convective heat transfer, a power law relates the Rayleigh number ($Ra$) to the Nusselt number ($Nu$) ~\cite{hardee1976boundary,merkin1979free}:
\begin{equation}
Nu_{conv} = 0.565 * Ra^{0.5}\,.
\label{eq:nu-conv}
\end{equation}

The Nusselt number is defined as: 
\begin{equation}
Nu = \frac{hD}{k_e}  = \frac{qD}{\Delta T \lambda}\,,
\end{equation}
where $h$ is the convective heat transfer coefficient (\si{\watt\per\meter\squared\kelvin}), $D$ is the characteristic length (burial depth), $\lambda$ is the thermal conductivity (\si{\watt\per\meter-\kelvin}), $q$ is the heat flux from the cable (\si{\watt}), and $\Delta T$ is the temperature difference between the cable surface and the ambient temperature. The Nusselt number for conduction ($Nu_{cond}$) is a function of the burial depth:
\begin{equation}
    Nu_{cond} = \frac{2}{\ln \left(\frac{4 D}{d}\right)}\,,
    \label{eq:nu-cond}
\end{equation}
where $D$ is the depth of burial and $d$ is the diameter of the cable.

The Nusselt number in the simulations are calculated as: 
\begin{equation}
Nu = \frac{hD}{\lambda} =  \frac{\frac{\partial (T_c - T)}{\partial y}|_{y=0}}{T_s -T_\infty}\,,
\end{equation}
where $T_c$ is the cable temperature and $T_\infty$ is the temperature near the ground surface.

\Cref{fig:ra-nu} shows the relationship between the Nusselt number and the Rayleigh number for different temperatures. The Nusselt number based on the simulation matches the Nusselt number for conduction (\cref{eq:nu-cond}) for low permeability conditions ($Ra < 1$). At high permeabilities ($Ra > 100$), the Nusselt number increases with the Rayleigh number as a power of 0.5 as in the convection relation in~\cref{eq:nu-conv}. A transition region exists for Ra between 1 and 100, where a power-law relation describes the combined influence of conduction and convection. In the transition zone, the simulation Nusselt number ($Nu_{sim}$) shows that the effect of conduction gradually decreases with increasing Rayleigh numbers, eventually reaching a fully convective flow and heat transfer.~\citeN{churchill1972general} expressed the heat transport process to vary uniformly between two limiting solutions of purely conductive and convective heat transfer:
\begin{equation}
    Nu = (Nu_{cond}^n + Nu_{conv}^n)^(1/n) \,,
\end{equation}
where $n$ is the closest integer value optimized for the best fit, in this study, we use $n = 5$.  The relationship between the heat transfer($Nu$) and flow conditions ($Ra$) shows that conduction dominates for soils with $Ra < 1$, convection is the dominant mechanism of heat transfer for $Ra > 100$, and a transition region where both conduction and convection heat transfer occurs for $Ra$ between 1 and 100.

\begin{figure}
    \centering
    \includegraphics{figs/ra-nu.eps}
    \caption{Variation of Nusselt number as a function of Rayleigh number for only conduction, only convection and heat transfer with conduction and convection.}
    \label{fig:ra-nu}
\end{figure}

\subsection*{Effect of burial depth}
Sediment mobility in submarine conditions may cause a variation in burial depth of up to 5 m in a year~\cite{young2001utilizing}. Changing the cable burial depth alters the distance between the cable and the overlying seawater heat sink, increasing the temperature gradient between the cable and the seawater interface. We vary the burial depth as 0.5, 1.0, and \SI{1.5}{\meter} for three different cable temperatures of \SI{70}{\celsius}, \SI{140}{\celsius}, and \SI{210}{\celsius}. The burial depth and temperature gradient have a linear relation with the Rayleigh number. By increasing the burial depth from 0.5 to \SI{1.0}{\meter}, we increase the Rayleigh number by a factor of 2. In contrast, by reducing the cable temperature from \SI{140}{\celsius} to \SI{70}{\celsius} the Rayleigh number halves. We evaluate the influence of burial depth and cable temperatures by comparing the behavior at the same Rayleigh numbers.

As the burial depth increases, the Rayleigh number increases, indicating that the deeper the burial is, the more likely the flow is convection.~\Cref{fig:ra-tnorm-depth} shows the heating efficiency for different Rayleigh numbers at different burial depths. Heating efficiency is the ratio of pipe temperature to cable temperature.~\Cref{eq:ra} shows that the burial depth influences the time-scale of conduction  ($t \propto D^2$) significantly more than the time-scale of convection ($t \propto 1/D$). In the low permeability condition dominated only by conduction (Ra < 1),~\cref{fig:ra-tnorm-depth} shows the normalized pipe temperature decreasing with the burial depth. However, this decrease in heat transfer efficiency is not significant, dropping by 2\% as burial depth increases from \SI{0.5}{\meter} to \SI{2}{\meter}.

As the Rayleigh number increases from 1 to 100, an increase in burial depth shows a marked decrease in the pipeline temperature. In this transition zone, where conduction and convection occur, the increase in burial depth drastically reduces the heat transfer efficiency. The combined effect of quadratic increase in the conduction time scale and the linear drop in the convective time scale deteriorates the heat transfer efficiency. When the permeability becomes more favorable to convective heat transfer ($Ra > 100$), the pipe temperature is almost negligible and remains indifferent to the cable burial depth.

We also observe that the burial depth, not the cable temperature, is the controlling factor in heat dissipation. At a given burial depth, the evolution of heating efficiency (normalized pipe temperature) follows the same curve irrespective of the cable temperature. For the range of burial depths, conductive heat transfer efficiency is not affected by the burial depth, while convective heat transfer shows a marked loss in efficiency with an increase in burial depth.

\begin{figure}
    \centering
    \includegraphics{figs/ra-tnorm-depth.eps}
    \caption{Effect of burial depth on pipe temperature.}
    \label{fig:ra-tnorm-depth}
\end{figure}

\section{Conclusions}

Deep-Sea pipelines are susceptible to hydrate formation at low temperatures. Inductive heating with piggyback cables reduces the risk of hydrate formation and improves the durability of the pipelines by minimizing corrosion. Inductive heating with piggyback cables loses a portion of the generated heat to the surrounding soil. The amount of heat lost to the surrounding environment depends on many parameters: soil conductivity, permeability, cable temperature, and burial depth. The current practice of calculating the current rating required to maintain the cable temperature only considers conductive heat transfer, neglecting the convective heat dissipation.

We study the different modes of heat transfer for pipelines with inductive heating using 2D Finite Difference analysis. Low permeability soil only experiences conductive heat transfer showing a uniform symmetric heat distribution, thus losing minimum heat to the surrounding soil;  Whereas highly permeable soils dissipate a significant portion of the heat to the surrounding soil, barely heating the pipeline due to convective heat transfer with buoyancy forces driving the heat away from the pipeline and towards the ground surface. A transition region exists where both conduction and convection occur. The Rayleigh number, not just permeability, is the controlling factor influencing the transition in the heat transfer behavior. We identify a critical Rayleigh-Darcy number of 1 as the controlling value separating conduction from convection, much below the established boundary of $Ra = 12$ in the literature for rectangular domains with a constant heat source. At $Ra < 1$, conduction dominates the heat dissipation from the cables and is the ideal zone for inductive heating of pipelines. As Ra increases in the range of 1 - 100, both conduction and convection occur, and heating efficiency significantly drops from 73.5\% to 13\%. When $Ra > 100$, almost no pipeline heating occurs for high permeability soils, losing all the heat to the ground above. For low permeability soil conditions, the depth of burial has a negligible effect on the heating efficiency. Whereas for soils with high permeabilities, an increase in burial depth significantly decreases the heating efficiency.
\pagebreak
%
\appendix
%
\section{Notation}
\label{app:notation}
\nopagebreak
\begin{tabular}{r  @{\hspace{1em}=\hspace{1em}}  l}
$\alpha$    & Thermal diffusivity (\si{\meter\squared\per\second})\\
$\beta$     & Thermal expansion of water (\si{\per\kelvin}); \\
$\gamma$    & Unit weight (\si{\kilo\newton\per\meter\cubed});\\
$\lambda$   & Thermal conductivity (\si{\watt\per\meter-\kelvin}); \\
$\mu$       & Dynamic viscosity of fluid (\si{\pascal\cdot\second}); \\
$\rho$      & Mass density (\si{\kilogram\per\meter\cubed}); \\
$\nabla$    & Gradient; \\

$c$         & Courant–Friedrichs–Lewy number; \\
$c_{p}$     & Specific heat capacity (\si{\joule\per\kilo\gram\per\celsius}); \\
$d$         & Cable diameter (\si{\meter}); \\
$d_m$       & Mean grain diameter (\si{\meter}); \\
$e$         & Void ratio; \\
$D$         & Depth of burial (\si{\meter}); \\
$g$         & Acceleration due to gravity (\si{\meter\per\second\squared});\\
$G_s$       & Specific gravity of solid; \\
$h$         & Convective heat transfer coefficient (\si{\watt\per\meter\squared\kelvin}); \\
$k$         & Intrinsic permeability (\si{\meter\squared}); \\
$n$         & Porosity; \\
$Nu$        & Nusselt number; \\
$q$         & Heat flux (\si{\watt}); \\
$Ra^\prime$ & Rayleigh number; \\
$Ra$        & Rayleigh-Darcy number; \\
$t$         & time step (\si{\second}); \\
$\Delta x$   & Mesh size (\si{\meter}); \\
$T$         & Temperature (\si{\celsius}); \\
$\mathbf{u}$ & Fluid velocity (\si{\meter\per\second}); \\
\end{tabular}

\subsection{Subscripts}
\begin{tabular}{r  @{\hspace{1em}=\hspace{1em}}  l}
$_0$        & Initial state; \\
$_b$        & Bulk property; \\
$_f$        & Fluid property; \\
$_p$        & Pipe property; \\
$_s$        & Solid property; \\
$_{sat}$    & Saturated property; \\
$_w$        & Water property; \
\end{tabular}

\label{section:references}
\bibliography{ascexmpl-new}
